Modelling fine-scale correlation structures using hidden Markov models

### 2020-07-19 Report/Sections/models.tex
Updates from Overleaf

Traditional hidden Markov models (HMMs) are useful tools to \textit{blah blah blah - quick (5-7words) restatement of the main advantage of HMM}, and will be used as the core structure of our models.
However, traditional HMMs assume conditional independence between observations given the state sequence, and thus do not hold when the observations exhibit certain forms of significant correlation in time. Traditional HMMs model process autocorrelation with a single Markov chain, however many processes have temporal dependencies acting at different scales. \textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things. You can use some of the text I removed from below}. 
To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM: the conditionally auto-regressive hidden Markov model (CarHMM), hierarchical hidden Markov model (HHMM), and HMM-DFT. We will show how each of these variations can be altered and combined to form a wide variety of new models.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The HMM with discrete Fourier transform (HMM-DFT)}
\label{subsec:STFT}

The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM. In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform. For simplicity, we assume that the length of the fine-scale processes is constant (i.e. that $T^*_t = T^*$), although this need not be the case in general. Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour. We then suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right), \quad k = 0, 1, \ldots, T^*-1.
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example is as follows:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
In words, $z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure \ref{fig:HMM-DFT} shows a graphical representation of the HMM-DFT.

Once $z^*_t$ is calculated, it can be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. In total, the likelihood of the hierarchical HMM-DFT is as follows:
\begin{equation}
    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
    \label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

@@

\subsection{General hierarchical structures}

@@

Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, state-switching OU, HHMM, and HMM-DFT). These five models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs. In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models constructed from these building blocks.

>>>>

Traditional hidden Markov models (HMMs) are useful tools to \textit{blah blah blah - quick (5-7words) restatement of the main advantage of HMM}, and will be used as the core structure of our models.
However, traditional HMMs assume conditional independence between observations given the state sequence, and thus do not hold when the observations exhibit certain forms of significant correlation in time. Traditional HMMs model process autocorrelation with a single Markov chain, however many processes have temporal dependencies acting at different scales. \textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things. You can use some of the text I removed from below}. 
To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM: the conditionally auto-regressive hidden Markov model (CarHMM), hierarchical hidden Markov model (HHMM), and HMM with discrete Fourier transform (HMM-DFT). We will show how each of these variations can be altered and combined to form a wide variety of new models.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Accounting for periodicity with HMM-DFT}
\label{subsec:STFT}

The HMM-DFT incorporates hierarchical structure into an HMM differently than an HHMM (\textit{ref Fig 1}). In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform. For simplicity, we assume that the length of the fine-scale processes is constant (i.e. that $T^*_t = T^*$), although this need not be the case. Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour. We then suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right), \quad k = 0, 1, \ldots, T^*-1.
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example is as follows:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
In words, $z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure \ref{fig:HMM-DFT} represents the structure of the HMM-DFT.

Once $z^*_t$ is calculated, it can be treated as an observation of the HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. The likelihood of the HMM-DFT is as follows:
\begin{equation}
    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
    \label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

@@

\subsection{General structure for building complexity}

@@

Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, state-switching OU, HHMM, and HMM-DFT). These alternative models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs. In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models constructed from these building blocks.

### 2020-07-19 Report/Sections/results.tex
Updates from Overleaf

The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada. Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale as well as the distribution of accelerometer data within each behavioral state is import to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

@@

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

@@

It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

>>>>

The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada. Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale, as well as the distribution of accelerometer data within each behavioral state, is important to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

@@

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time periods were ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed dives within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

@@

Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
